Semantic Novelty Detection with Graphical Models

### 2011-05-06 semantic-nd/article.tex
More stuff on novelty detection.

Such an ordering can be defined by a R-value function on the input space. And then a threshold can be trained
on that function in order to accept an expected amount of cases at the cost of failing in others.

Based on bayes rule:

\begin{equation}
P(\overline{novel}|x) = \frac{P(x|\overline{novel}) P(\overline{novel})}{P(x)}
\end{equation}

@@

\subsection{Using Unlabelled Data}
Nonetheless in several cases its often the case that there is access to large amounts of unlabelled data.
Under that it becomes possible to obtain a better approach to the unconditional probability distribution than
the uniform one assumed in \autoref{sec:uniform-unconditional}.

For interesting and practical reasons it was assumed that the unconditional distribution
could be modelled with all the variables independent from each other.
Which translates as $P(x)=\prod{P_{x_i}(x_i)}$.
And so from the unlabelled data we only need to estimate $P_{x_i}$.
\autoref{fig:semisupervised-threshold} illustrates this in terms of graphs.

@@

Once again the novelty threshold would be given by $P_G(x)/P_U(x)$. This ratio between the two graphs
can be understood as compensating for an existing bias on the unconditional distribution.
Which is an important step to achieve a correct ordering of the input space for novelty thresholding.
This way our unconditional function perceives which features are slightly more biased to a certain value
and will compensate on the novelty calculation.

@@

And although it serves as a method for detecting new classes in the input, it gives us no control in what
variables novelty is being measured.
This is the novelty can come from any variable present on the graphs and we have no way to measure it.

@@

\begin{figure}
\caption{\label{fig:measure-room-novelty}In this case the only difference between $G$ and $G'$ is the room category variable,
         that only exists in the later.}
\end{figure}

From now on we refer to the measure $P_G(x)/P_{G'}(x)$, when $G$ and $G'$ as a novelty measure on the variable $R_{category}$.
As it attempts to measure on how much adding the variable $R_{category}$ helps to explain the given distribution of values.

Expanding the calculation of $P_G(x)/P_{G'}(x)$ with the sum-product technique. Leads to:

@@

>>>>

Using bayes rule on the reversed order $P(\overline{novel}|x)$ and assuming a constant $P(\overline{novel})$
we obtain a ratio between a conditional and unconditional probabilites on the input $x$.
Such a ratio is a suitable function for implementing an optimal novelty detector system with
thresholding.

\begin{equation}
\label{eq:novelty-threshold}
P(\overline{novel}|x) = \frac{P(x|\overline{novel}) P(\overline{novel})}{P(x)} = k\frac{P(x|\overline{novel})}{P(x)}
\end{equation}

@@

Here $P_{U}(x)$ can be seen as a normalizing factor to lever all the $P_G(x)$ on any set of variables $x$ into the same
measure units, such that a static threshold can be implemented.
For example in very large sets of variables $x$, a large dimension space, the conditional probability would spread over
the tails of the functions and would always yield very small values.
As so a novelty measure is seen as a ratio on how much introducing the known concepts helps to understand the observed result.

\subsection{Using Unlabelled Data}
Nonetheless its often the case that there is access to large amounts of unlabelled data.
Under that it becomes possible to obtain a better approach to the unconditional probability distribution than
the uniform one assumed in \autoref{sec:uniform-unconditional}.

For interesting and practical reasons it was assumed that the unconditional distribution
could be modelled with all the variables independent from each other.
Which translates as $P(x)=\prod{P_{x_i}(x_i)}$.
And so from the unlabelled data we only need to estimate $P_{x_i}$.
\autoref{fig:semisupervised-threshold} illustrates a graph $I$ to model independent variables.

@@

Once again the novelty threshold would be given by $P_G(x)/P_I(x)$.
This time the addition of the variables unconditional factors can be understood as an attempt to compensate
for an existing bias on the unconditional distribution.
Which is an important step to achieve a correct ordering of the input space for novelty thresholding.
This way our unconditional function perceives which features are slightly more biased to a certain value
and will compensate on the novelty calculation.

@@

And although it serves as a method for detecting new classes in the input, it gives us no control in what
is novel in the input.
Recall that the graph $G$ used by Dora gives us a lot of semantic and conceptual information that we are
interested in analizying and improving with time. Being able to measure if the novel case derives from
a novel room category or from any other variables would be highly desirable on the system.

@@

\begin{figure}[h]
\subfloat[Conditional probability distribution $G$]{\includegraphics[width=0.45\textwidth]{figures/simple-conditional-prob-graph.pdf}}
\hfill
\subfloat[Unconditional probability distribution $G'$]{\includegraphics[width=0.45\textwidth]{figures/simple-unconditional-prob-graph.pdf}}

\caption{\label{fig:measure-room-novelty}In this case the only difference between $G$ and $G'$ is the room category variable
         and the associated factors that only exists in the former.}
\end{figure}

The notation $N_v$ is now used to refer to the measure $P_G(x)/P_{G'}(x)$, when $G$ and $G'$ differ only in the variable node $v$ and
its associated factor nodes.
And use it as a novelty measure of the variable $v$, as it attempts to measure on how much adding the variable $v$ helps to explain the
given distribution of values.

By assuming the graph is a tree and blabla, and expanding the calculation of $P_G(x)/P_{G'}(x)$ with the sum-product technique, leads to:

\begin{equation}
N_v(x) = \frac{\sum_{s \in States(v)}x / normalizing_{factor1}}{\prod{...} / normalizing_{factor2}}
\end{equation}

blablablaaa....
I haven't done the math here yet.... but it should give us a method for measuring on how much each variable is contributing to the novelty
threshold.
Also due to the nature of the graphical models it should be possible to calculate in $O(N)$ time for all the $N$ variables.
After the whole variables can be sorted by the contributions and presented to the user explaining where the novelty comes from.

Quotting Andrzej: """ Being able to automatically explain why a class is novel and what
makes it different from any of the other classes -- This is super important for communication with the user and active learning of new room categories."""

@@

In order to verify the improvements obtained by using unlabelled data, a fictitious simple dataset was generated.
The fictitious distribution assumes that an independent and variable size set of features $x$ is generated by a
given room category category. In whole there was 11 different room categories and 7 different measured properties types.

The known data was extracted by generating samples from 6 of the 11 classes. Together with the generated known data, the
room category that generated is given as label. Using that we built the factors for the generation of graphs representing
the conditional distribution $G$.
Using unlabelled data we trained also a graph $I$ representing an unconditional distribution where all features are independent.

\begin{figure}[h!]
\centering
\includegraphics[width=0.60\textwidth]{figures/roc-simple.pdf}

This graph is from a previous experience. Where there was even no need for the normalizing factor.

\caption{\label{fig:fictitious-roc}ROC curve comparing novelty detection performance between an optimal detector,
         $P_G(x)/P_U(x)$ a graph ratio considering an uniform unconditional distribution and
         $P_G(x)/P_I(x)$ a graph ratio considering an independent unconditional distribution}
\end{figure}

Since the distribution is fictitious we were able to create an optimal novelty detector, as we can exactly measure $P(novel|x)$.
That served as a comparison on how optimal a detector could be.
A ROC curve was plotted for the three methods on \autoref{fig:fictitious-roc}.
As possible to see a correct estimation on the unconditional probability is important for obtaining a correct ordering of the
input space.
